---
title: "Compare reference genotypes for vireo"
author: "Ariel Hippen"
date: '2022-06-27'
output: 
  pdf_document:
    toc: yes
    toc_depth: '4'
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
suppressPackageStartupMessages({
  library(data.table)
  library(scater)
  library(ggplot2)
  library(dplyr)
  library(caret)
  library(yaml)
})
```



## Load data

```{r options}
# We have two sets of pooled samples, named after the date they were run (12162021 and 01132022).
sample_id <- "12162021"

params <- read_yaml("../../config.yml")
data_path <- params$data_path
local_data_path <- params$local_data_path
samples <- params$samples
```

```{r}
# Load hash and vireo (genetic) demultiplexing assignments
# Note: variable data_path is loaded from config.R
hashing <- fread(paste(data_path, "pooled_tumors", sample_id,
                       "Cellranger/outs/multi/multiplexing_analysis/assignment_confidence_table.csv", sep = "/"))

vireo_path <- paste(data_path, "pooled_tumors", sample_id, "vireo", sep = "/")
chunk_ribo <- fread(paste(vireo_path, "chunk_ribo/donor_ids.tsv", sep = "/"))
dissociated_ribo <- fread(paste(vireo_path, "dissociated_ribo/donor_ids.tsv", sep = "/"))
dissociated_polyA <- fread(paste(vireo_path, "dissociated_polyA/donor_ids.tsv", sep = "/"))
```

```{r}
# Load SingleCellExperiment object for plotting
sce <- readRDS(paste("../../data/sce_objects/", sample_id, ".rds", sep = ""))
```

```{r}
# Filter hashing and vireo to cells in sce object, all others failed miQC filtering
hashing <- subset(hashing, hashing$Barcodes %in% sce$Barcode)
chunk_ribo <- subset(chunk_ribo, chunk_ribo$cell %in% sce$Barcode)
dissociated_ribo <- subset(dissociated_ribo, dissociated_ribo$cell %in% sce$Barcode)
dissociated_polyA <- subset(dissociated_polyA, dissociated_polyA$cell %in% sce$Barcode)
```

```{r}
# Subset down and join into one matrix
setnames(chunk_ribo, "donor_id", "chunk_ribo_assignment")
setnames(dissociated_ribo, "donor_id", "dissociated_ribo_assignment")
setnames(dissociated_polyA, "donor_id", "dissociated_polyA_assignment")

assignments <- full_join(subset(chunk_ribo, select = c("cell", "chunk_ribo_assignment")),
                         subset(dissociated_ribo, select = c("cell", "dissociated_ribo_assignment"))) %>% 
               full_join(., subset(dissociated_polyA, select = c("cell", "dissociated_polyA_assignment")))
assignments$chunk_ribo_assignment <- as.factor(assignments$chunk_ribo_assignment)
assignments$dissociated_ribo_assignment <- as.factor(assignments$dissociated_ribo_assignment)
assignments$dissociated_polyA_assignment <- as.factor(assignments$dissociated_polyA_assignment)
```

## Compare vireo assignments

```{r}
# Run confusion matrix
confusionMatrix(assignments$chunk_ribo_assignment, assignments$dissociated_ribo_assignment)$table
confusionMatrix(assignments$chunk_ribo_assignment, assignments$dissociated_polyA_assignment)$table
confusionMatrix(assignments$dissociated_ribo_assignment, assignments$dissociated_polyA_assignment)$table
```

There seems to be a lot of concordance across the sets, which is encouraging. I want to check if there are patterns in the off-diagonal cells.

Since donor labels are assigned randomly, all "donor2" cells in chunk_ribo might be called "donor0" in dissociated_ribo. I'll have to manually reassign them to a new name. I'll use A, B, C, and D.

```{r}
assignments$CR <- assignments$chunk_ribo_assignment
assignments$CR <- recode(assignments$CR,
                         "donor0" = "A",
                         "donor1" = "B",
                         "donor2" = "C",
                         "donor3" = "D")

assignments$DR <- assignments$dissociated_ribo_assignment
assignments$DR <- recode(assignments$DR,
                         "donor0" = "D",
                         "donor1" = "A",
                         "donor2" = "B",
                         "donor3" = "C")

assignments$DP <- assignments$dissociated_polyA_assignment
assignments$DP <- recode(assignments$DP,
                         "donor0" = "C",
                         "donor1" = "A",
                         "donor2" = "B",
                         "donor3" = "D")
```

```{r}
# Combine "doublet" and "unassigned" into one category, the distinction doesn't seem important
assignments[assignments$CR == "doublet", ]$CR <- "unassigned"
assignments[assignments$DR == "doublet", ]$DR <- "unassigned"
assignments[assignments$DP == "doublet", ]$DP <- "unassigned"

# Redo factor levels
assignments$CR <- as.factor(as.character(assignments$CR))
assignments$DR <- as.factor(as.character(assignments$DR))
assignments$DP <- as.factor(as.character(assignments$DP))
```

```{r}
# Look for patterns in cells where there are disagreements
assignments$disagreement <- ifelse(assignments$CR != assignments$DR |
                                     assignments$CR != assignments$DP,
                                   1, 0)
sce$disagreement <- assignments$disagreement
table(sce$disagreement)
```
```{r}
plotUMAP(sce, colour_by = "disagreement")
```

The cells with "disagreements" seem to be more or less randomly distributed. I'm happy with this.

## Compare to hash demultiplexing

One last question: these are extremely concordant, but does one bulk type give a marginal increase in concordance with hash demultiplexing?

```{r}
#Add hashing results to assignments table
hashing <- subset(hashing, select = c("Barcodes", "Assignment"))
setnames(hashing, c("cell", "hash_assignment"))
assignments <- full_join(hashing, assignments)
```
```{r}
# Switch hash results to A/B/C/D notation
assignments$H <- assignments$hash_assignment
assignments[assignments$H == "Blanks" | assignments$H == "Multiplet" | assignments$H == "Unassigned", ]$H <- "unassigned"
assignments$H <- as.factor(assignments$H)
assignments$H <- recode(assignments$H,
                         "anti-human_Hashtag1" = "D",
                         "anti-human_Hashtag2" = "C",
                         "anti-human_Hashtag3" = "A",
                         "anti-human_Hashtag4" = "B")
```

```{r}
# Count number of disagreements between hashing and each run of vireo
length(which(assignments$H != assignments$CR))
length(which(assignments$H != assignments$DR))
length(which(assignments$H != assignments$DP))

confusionMatrix(assignments$H, assignments$CR)$table
confusionMatrix(assignments$H, assignments$DR)$table
confusionMatrix(assignments$H, assignments$DP)$table
```

Conclusion: there is only the tiniest of differences in the number of cells that are picked up by genetic demultiplexing but not hash demultiplexing, and literally no difference in the number of cells that get confidently assigned by both. Looks like it doesn't matter.
