---
title: "Manual Annotation of Pooled Cells"
author: "Ariel Hippen"
date: '2022-06-14'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is the code I used to cluster the pooled data and identify markers that eventually turned into cell types. Most of the heavy lifting was done with internet searches of the markers findMarkers() recovered. Notes from that search are stored in a google doc in my drive; I'm currently trying to decide the best way to preserve that info for reproducibility.

```{r packages}

suppressPackageStartupMessages({
  library(data.table)
  library(scater)
  library(scran)
  library(batchelor)
  library(igraph)
  library(WebGestaltR)
})

```

## Load and prep data

```{r load}
# Load and combine pooled data objects
pool1 <- readRDS("../../data/sce_objects/12162021.rds")
pool1$Pool <- "12162021"
pool2 <- readRDS("../../data/sce_objects/01132022.rds")
pool2$Pool <- "01132022"

sce <- cbind(pool1, pool2)
```

Plot and check for pool-specific structure suggesting batch effects

```{r}
set.seed(531)
sce <- runUMAP(sce,
               BNPARAM = BiocNeighbors::AnnoyParam(),
               BPPARAM = BiocParallel::MulticoreParam(),
               min_dist = 0.5,  repulsion_strength = 0.25,
               spread = 0.7,
               n_neighbors = 15)
plotUMAP(sce, colour_by = "Pool")
```

Run batchelor to correct for batch effects

```{r}
set.seed(531)
mnn <- fastMNN(sce,  batch = sce$Pool,
               BSPARAM = BiocSingular::IrlbaParam(deferred = TRUE),
               BNPARAM = BiocNeighbors::AnnoyParam(),
               BPPARAM = BiocParallel::MulticoreParam())

reducedDim(sce,  "MNN") <- reducedDim(mnn,  "corrected")
rm(mnn)
```

Now we can replot to confirm that the pool-specific structure looks better

```{r}
set.seed(531)
sce <- runUMAP(sce,
               dimred = "MNN",
               BNPARAM = BiocNeighbors::AnnoyParam(),
               BPPARAM = BiocParallel::MulticoreParam(),
               min_dist = 0.5,  repulsion_strength = 0.25,
               spread = 0.7,
               n_neighbors = 15)
plotUMAP(sce, colour_by = "Pool")
```

## Check for basic markers

Look for basic cell type markers, PTPRC (CD45) for immune, ACTA2 (smooth muscle actin) for fibroblasts, and EPCAM and PAX8 for epithelial/cancer cells.

```{r}
plotUMAP(sce, colour_by = "PTPRC")
plotUMAP(sce, colour_by = "EPCAM")
plotUMAP(sce, colour_by = "PAX8")
plotUMAP(sce, colour_by = "ACTA2")
```

## Cluster cells

Creating the SNN graph takes >1hr, so I will leave the code in to generate the clusters but after the first time just load in the existing RDS object. Also, I initially ran this several times with k=30, 50, and 80, but decided to go with the resolution at 50.
```{r}
#g <- buildSNNGraph(sce,
#                   k = 50,   # higher = bigger clusters
#                   BNPARAM = BiocNeighbors::AnnoyParam(),
#                   BPPARAM = BiocParallel::MulticoreParam())
#clusters <- as.factor(igraph::cluster_louvain(g)$membership)
#sce$clusters <- clusters
#saveRDS(sce,  file = "../../data/sce_objects/pooled_clustered_50.rds")
sce <- readRDS("../../data/sce_objects/pooled_clustered_50.rds")

plotUMAP(sce, colour_by = "clusters")
```

## Annotate clusters

For the annotation, we'll be using scran's findMarkers function. It returns results for all genes regardless of significance, so for clarity we'll pull out the ones with a FDR of <=0.05.
```{r}
getOnlySignificantGenes <- function(mylist) {
  for (i in 1:length(mylist)) {
    x <- mylist[[i]]
    x <- subset(x, select = c("p.value", "FDR"))
    y <- subset(x, x$FDR <= 0.05)
    if (nrow(y) > 0) {
      y <- y[order(y$FDR), ]
    }
    mylist[[i]] <- y
  }
  as.list(mylist)
}
```

Let's start by seeing if we can pin down any particular clusters with genes that are upregulated compared to all other clusters
```{r}
markers <- findMarkers(sce,  groups = sce$clusters,
                       block = sce$Pool,  # use to get within-donor DE
                       direction = "up",  lfc = 1.5,
                       pval.type = "all",
                       BPPARAM = BiocParallel::MulticoreParam())

sig_genes <- getOnlySignificantGenes(markers)
sig_genes
```

Still looking for global markers, but with a relaxed lfc
```{r}
markers <- findMarkers(sce,  groups = sce$clusters,
                       block = sce$Pool,  # use to get within-donor DE
                       direction = "up",  lfc = 1.3,
                       pval.type = "all",
                       BPPARAM = BiocParallel::MulticoreParam())

sig_genes <- getOnlySignificantGenes(markers)
sig_genes
```

Look for markers that are upregulated compared to at least some clusters

```{r}
markers <- findMarkers(sce,  groups = sce$clusters,
                       block = sce$Pool,  # use to get within-donor DE
                       direction = "up",  lfc = 1.5,
                       pval.type = "some",
                       BPPARAM = BiocParallel::MulticoreParam())

sig_genes <- getOnlySignificantGenes(markers)
sig_genes
```

Let's try running WebGestaltR, which runs overrepresentation analysis
```{r, warning=FALSE}
reference_genes <- rownames(sce)
for (i in 1:length(unique(sce$clusters))) {
  cluster_genes <- sig_genes[[i]]
  pathways <- WebGestaltR(enrichDatabase = "geneontology_Biological_Process",
                                  interestGene = rownames(cluster_genes),
                                  interestGeneType = "genesymbol",
                                  referenceGene = reference_genes,
                                  referenceGeneType = "genesymbol",
                                  isOutput = FALSE)
  pathways$userId <- NULL; pathways$overlapId <- NULL

  print(head(pathways))
}
```


After some grueling googling, I have a tentative identification for most clusters. I'm going to try breaking the larger groups up by these definitions and seeing what happens.

### Fibroblasts
```{r}
#fibroblasts <- sce[, sce$clusters %in% c(2, 6, 10)]
#g <- buildSNNGraph(fibroblasts,
#                   k = 50,   # higher = bigger clusters
#                   BNPARAM = BiocNeighbors::AnnoyParam(),
#                   BPPARAM = BiocParallel::MulticoreParam())
#clusters <- as.factor(igraph::cluster_louvain(g)$membership)
#fibroblasts$clusters <- clusters
#saveRDS(fibroblasts,  file = "../../data/sce_objects/fibroblasts.rds")
fibroblasts <- readRDS("../../data/sce_objects/fibroblasts.rds")

plotUMAP(fibroblasts, colour_by = "clusters")
```

```{r}
markers <- findMarkers(fibroblasts,  groups = fibroblasts$clusters,
                       block = fibroblasts$Pool,  # use to get within-donor DE
                       direction = "up",  lfc = 1.5,
                       pval.type = "some",
                       BPPARAM = BiocParallel::MulticoreParam())

sig_genes <- getOnlySignificantGenes(markers)
sig_genes
```

### Immune cells
```{r}
# Looking at the somewhat enigmatic immune clusters (5 was clearly macrophages)
#immune <- sce[, sce$clusters %in% c(1, 3, 8, 11, 12)]
#g <- buildSNNGraph(immune,
#                   k = 50,   # higher = bigger clusters
#                   BNPARAM = BiocNeighbors::AnnoyParam(),
#                   BPPARAM = BiocParallel::MulticoreParam())
#clusters <- as.factor(igraph::cluster_louvain(g)$membership)
#immune$clusters <- clusters
#saveRDS(immune,  file = "../../data/sce_objects/immune.rds")
immune <- readRDS("../../data/sce_objects/immune.rds")

plotUMAP(immune, colour_by = "clusters")
```

```{r}
markers <- findMarkers(immune,  groups = immune$clusters,
                       block = immune$Pool,  # use to get within-donor DE
                       direction = "up",  lfc = 1.5,
                       pval.type = "some",
                       BPPARAM = BiocParallel::MulticoreParam())

sig_genes <- getOnlySignificantGenes(markers)
sig_genes
```

### APCs
```{r}
apcs <- immune[, immune$clusters %in% c(5, 8)]
markers <- findMarkers(apcs,  groups = apcs$clusters,
                       block = apcs$Pool,  # use to get within-donor DE
                       direction = "up",  lfc = 1.5,
                       pval.type = "all",
                       BPPARAM = BiocParallel::MulticoreParam())

sig_genes <- getOnlySignificantGenes(markers)
sig_genes
```

### T cells
```{r}
tcells <- sce[, sce$clusters %in% c(1, 3, 12)]
markers <- findMarkers(tcells,  groups = tcells$clusters,
                       block = tcells$Pool,  # use to get within-donor DE
                       direction = "up",  lfc = 1.5,
                       pval.type = "some",
                       BPPARAM = BiocParallel::MulticoreParam())
sig_genes <- getOnlySignificantGenes(markers)
sig_genes
```

